---
title: "Create zooplankton database"
author: "Camille"
date: "6/8/2020"
output: html_document
toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose

The purpose of this document is to aggregate the individual zooplankton datasets into one comprehensive database.  

## Load libraries

The necessary libraries are loaded.  These include a tidy way to read in CSV files, a package for streamlined data manipulation, and a package for date manipulation.

```{r load_libraries, message=FALSE}
# -------- Load libraries --------
library(readr)
library(dplyr)
library(lubridate)
```

## Initialize filepaths

File paths for the data layers are then initialized, based off of the base data filepath.  

```{r init_fps}
# -------- Initialize filepaths --------
# Data filepath
fp_data <- "./calanus_data/Data"
# ECOMON filepath
fp_ecomon <- file.path(fp_data, "ECOMON")
# Runge filepath
fp_runge <- file.path(fp_data, "Runge")
# CPR filepath
fp_cpr <- file.path(fp_data, "CPR")
# Databases filepath
fp_db <- file.path(fp_data, "Databases")
```

## Load ECOMON data

The ECOMON data is then loaded.  The centropage, C. finmarchicus, and pseudocalanus fields are renamed so they match the other datasets.  The date field is converted into year, month, and day.  The time field is converted into hours and minutes.  Empty fields are added for the zooplankton stages that were not included in the ECOMON data, but were included in other datasets.  Then, the relevant fields are selected from the dataset.  

```{r load_ecomon, warning=FALSE, message=FALSE}
# -------- Load ECOMON data ---------
# Read in data -- version 3.5
ecomon_dat <- readr::read_csv(file.path(fp_ecomon, "EcoMon_Plankton_Data_v3_5.csv")) %>%
  # Rename calanus column for consistency across datasets
  dplyr::rename(ctyp_adult = ctyp_10m2, # Centropages
                cfin_CV_CVI = calfin_10m2, # C. finmarchicus
                pseudo_adult = pseudo_10m2) %>% # Pseudocalanus
  # Break up date and time into individual units
  # Add missing columns for consistency across datasets
  dplyr::mutate(year = lubridate::year(as.Date(date, format = '%d-%b-%y')),
                month = lubridate::month(as.Date(date, format = '%d-%b-%y')),
                day = lubridate::day(as.Date(date, format = '%d-%b-%y')),
                hour = lubridate::hour(lubridate::hms(time)),
                minute = lubridate::minute(lubridate::hms(time)),
                cfin_CI = NA,
                cfin_CII = NA,
                cfin_CIII = NA,
                cfin_CV = NA,
                cfin_CIV = NA,
                cfin_CVI = NA,
                cfin_CI_CIV = NA,
                ctyp_CIII = NA,
                ctyp_CIV = NA, 
                ctyp_CV = NA,
                ctyp_C = NA, 
                ctyp_CIV_CV = NA, 
                ctyp_CIV_CVI = NA,
                ctyp_CVI = NA,
                ctyp_CV_CVI = NA, 
                psuedo_CII = NA,
                pseudo_CV = NA, 
                pseudo_C = NA, 
                pseudo_CVI = NA, 
                pseudo_CI_CIV = NA,
                dataset = "ECOMON") %>%
  # Select relevant columns
  dplyr::select(station, year, month, day, 
                hour, minute, lat, lon,
                cfin_CI, cfin_CII, cfin_CIII,
                cfin_CIV, cfin_CVI, cfin_CV,
                cfin_CI_CIV, cfin_CV_CVI, ctyp_adult,
                ctyp_CIII, ctyp_CIV, ctyp_CV,
                ctyp_C, ctyp_CIV_CV, ctyp_CIV_CVI,
                ctyp_CVI, ctyp_CV_CVI, pseudo_adult, 
                psuedo_CII, pseudo_CV, pseudo_C, 
                pseudo_CVI, pseudo_CI_CIV, dataset)
  
```

## Load Runge data

The Runge data is loaded  The year, month, day, latitude, and longitude fields and the fields for the different life stages of C. finmarchicus are selected and renamed so they match the other datasets.  Empty fields are added for the zooplankton species stages that were not included in the Runge data, but were included in other datasets.  

```{r load_runge, warning=FALSE, message=FALSE}
# -------- Load Runge data --------
# Read in data
runge_dat <- readr::read_csv(file.path(fp_runge, "UBER_ZOOBIO_2012_Aug16_2017_edit4GAMM.csv")) %>%
  # Add empty station, hour, and minute fields
  dplyr::mutate(station = NA,
                hour = NA, 
                minute = NA) %>%
  # Select relevant columns
  dplyr::select(station, YEAR, MNTH, DAY, hour, minute, `Lat Decimal`, `Log Decimal`, 
                `Calanus_finmarchicusCI(M2)`,
                `Calanus_finmarchicusCII(M2)`,
                `Calanus_finmarchicusCIII(M2)`,
                `Calanus_finmarchicusCIV(M2)`,
                `Calanus_finmarchicusCV(M2)`) %>%
  # Rename columns to match other datasets
  dplyr::rename(cfin_CI = `Calanus_finmarchicusCI(M2)`,
                cfin_CII = `Calanus_finmarchicusCII(M2)`,
                cfin_CIII = `Calanus_finmarchicusCIII(M2)`,
                cfin_CIV = `Calanus_finmarchicusCIV(M2)`,
                cfin_CV = `Calanus_finmarchicusCV(M2)`,
                year = YEAR,
                month = MNTH,
                day = DAY,
                lat = `Lat Decimal`,
                lon = `Log Decimal`) %>%
  # Add missing columns for consistency across datasets
  dplyr::mutate(cfin_CVI = NA,
                cfin_CI_CIV = NA,
                cfin_CV_CVI = NA,
                ctyp_adult = NA,
                ctyp_CIII = NA,
                ctyp_CIV = NA, 
                ctyp_CV = NA,
                ctyp_C = NA, 
                ctyp_CIV_CV = NA, 
                ctyp_CIV_CVI = NA,
                ctyp_CVI = NA,
                ctyp_CV_CVI = NA, 
                pseudo_adult = NA,
                psuedo_CII = NA,
                pseudo_CV = NA, 
                pseudo_C = NA, 
                pseudo_CVI = NA, 
                pseudo_CI_CIV = NA,
                dataset = "RUNGE")
```

## Load CPR data

The continuous plankton recorder (CPR) data is loaded.  Because the headers included two rows, the headers are read in first.  The first and second rows are then concatonated for each field so there is one row of headers.  The rest of the data is then read in, and the headers are appended back onto the data.  The relevant fields are selected and then renamed.  Because this dataset has all of the stages of the zooplankton species of interest, we did not have to add empty fields.    

```{r load_cpr, warning=FALSE, message=FALSE}
# -------- Load CPR data --------
# Read in headers
cpr_header <- readr::read_csv(file.path(fp_cpr, "GOMCPR_trim_cop.csv"), n_max = 2, col_names = FALSE)
# Concatonate fist and second row of headers
cpr_header <- paste0(cpr_header[1,], cpr_header[2,])
# Read in data
cpr_dat <- readr::read_csv(file.path(fp_cpr, "GOMCPR_trim_cop.csv"), skip = 2, col_names = FALSE)
# Reattach headers to cpr_data
names(cpr_dat) <- cpr_header
# Select relevant columns
cpr_dat <- cpr_dat %>% 
  dplyr::select(`'Station'NA`, `'Year'NA`, `'Month'NA`,
                `'Day'NA`, `'Hour'NA`, `'Minute'NA`, 
                `Latitude (degrees N)'NA`, `Longitude (degrees W)'NA`,
                contains("Calanus finmarchicus"),
                contains("Centropages typicus"),
                contains("'Pseudocalanus'")) %>%
  dplyr::mutate(dataset = "CPR")
# Rename all columns
names(cpr_dat) <- c("station", "year", "month",
                    "day", "hour", "minute", 
                    "lat", "lon", "cfin_CI", 
                    "cfin_CII", "cfin_CIII", 
                    "cfin_CIV", "cfin_CV", "cfin_CVI", 
                    "cfin_CI_CIV", "cfin_CV_CVI", "ctyp_adult",
                    "ctyp_CIII", "ctyp_CIV", "ctyp_CV",
                    "ctyp_C", "ctyp_CIV_CV", "ctyp_CIV_CVI",
                    "ctyp_CVI", "ctyp_CV_CVI", "pseudo_adult",
                    "psuedo_CII", "pseudo_CV", "pseudo_C", 
                    "pseudo_CVI", "pseudo_CI_CIV", "dataset")
```

## Create database

To create the comprehensive database, the three datasets are bound together row-wise.  Missing values (i.e. -9999) are set to 0.  Fields for the total numbers of each type of zooplankton are computed using rowSums().  Finally, the database is written to a CSV. 

```{r create_database}
# -------- Create database --------
# Combine individual datasets
database <- rbind(ecomon_dat, runge_dat, cpr_dat) 
# Replace missing values with 0
database[database == -9999] <- 0
# Add column for total C. finmarchicus across all stages
database$cfin_total <- rowSums(database %>% select(contains("calfin")), na.rm = TRUE)
# Add column for total centropages across all stages
database$ctyp_total <- rowSums(database %>% select(contains("ctyp")), na.rm = TRUE)
# Add column for total pseudocalanus across all stages
database$pseudo_total <- rowSums(database %>% select(contains("pseudo")), na.rm = TRUE)
# Write database to csv for future use
write_csv(database, file.path(fp_db, "zooplankton_database.csv"))
```


